---
title: "MR of protein on disease following PRS-protein assocs"
author: "Gibran Hemani"
execute:
    cache: true
---

Narrative

- Proteome-wide forward MR finds more assocs than restricting forward MR to those found from PRS
- There is very little agreement between cis and trans instruments
- Slightly more likely to find evidence from trans if there is evidence from cis
- significant cis MR agrees with PRS, no relationship between PRS and non-sig cis MR
- trans MR much stronger relationship with PRS, which is slightly attenuated after steiger filtering
- most instruments pass steiger filtering but all those that don't are trans
- 11 examples of bi-directional effects
- todo: coloc for cis-MR


```{r}
library(dplyr)
library(ieugwasr)
library(tidyr)
library(TwoSampleMR)
library(ggplot2)
library(here)
library(metafor)
library(readxl)
```

```{r}
traits <- read_xlsx(here("data", "Supplementary_Table5.xlsx"))
# gi <- ieugwasr::gwasinfo(traits$opengwasid)
gi <- readRDS(here("data", "gi.rds"))
load(here("data", "all.rdata"))
coloc_res <- readRDS(here("data", "cis_coloc_res.rds"))
coloc_res <- coloc_res %>%
    select(-c(id.outcome)) %>%
    left_join(., gi %>% select(trait, id.outcome = id), by=c("outcome"="trait")) %>%
    select(-c(outcome)) %>%
    left_join(., traits %>% select(outcome=code, id.outcome=opengwasid), by="id.outcome")

res <- res %>% select(-c(outcome)) %>%
    left_join(., traits %>% select(outcome=code, id.outcome=opengwasid), by="id.outcome")

res_cis <- res_cis %>% select(-c(outcome)) %>%
    left_join(., traits %>% select(outcome=code, id.outcome=opengwasid), by="id.outcome")

res_trans <- res_trans %>% select(-c(outcome)) %>%
    left_join(., traits %>% select(outcome=code, id.outcome=opengwasid), by="id.outcome")

res_trans_sf <- res_trans_sf %>% select(-c(outcome)) %>%
    left_join(., traits %>% select(outcome=code, id.outcome=opengwasid), by="id.outcome")

res_trans_sf_relaxed <- res_trans_sf_relaxed %>% select(-c(outcome)) %>%
    left_join(., traits %>% select(outcome=code, id.outcome=opengwasid), by="id.outcome")

table(coloc_res$outcome %in% traits$code)
table(coloc_res$outcome %in% res_cis$outcome)
table(coloc_res$outcome %in% res$outcome)
table(coloc_res$id.outcome %in% res$id.outcome)

coloc_res_s <- subset(coloc_res, PP.H4.abf >= 0.8)

res_cis_coloc <- subset(res_cis, paste(id.exposure, id.outcome) %in% paste(coloc_res_s$id.exposure, coloc_res_s$id.outcome))
```

```{r}
cis_mr <- left_join(
    res_cis,
    coloc_res %>% dplyr::select(id.exposure, id.outcome, H4=PP.H4.abf, H3=PP.H3.abf, leadsnp=snp),
    by=c("id.exposure", "id.outcome")
) %>% filter(p.adjust(pval, "fdr") < 0.05) %>%
arrange(desc(H4)) %>%
select(exposure, outcome, id.outcome, b, se, pval, H4, H3, leadsnp)
table(!is.na(cis_mr$H4))
head(cis_mr)
write.csv(cis_mr, file=here("results", "cis_mr.csv"))
```


Show MR results that have IVW / Wald ratio P val bonferroni corrected significant

```{r}
res %>% group_by(method) %>%
  summarise(n=n(), nsig = sum(fdr < 0.05, na.rm=T))
```

```{r}
ressig <- subset(res, fdr < 0.05)
dim(ressig)
```

Compare forward MR against reverse PRS results

```{r}
temp <- inner_join(subset(res, method %in% c("Inverse variance weighted", "Wald ratio")), prs_pairs, by=c("id.exposure"="prot", "id.outcome"="opengwasid"))
temp$mr_sig <- p.adjust(temp$pval.x, "fdr") < 0.05
ggplot(temp, aes(beta, b)) +
  geom_hline(yintercept=0, linetype="dotted") +
  geom_vline(xintercept=0, linetype="dotted") +
  geom_point(aes(colour=mr_sig)) +
  geom_smooth(method="lm") +
  labs(x="PRS disease -> protein", y="MR protein -> disease") +
  facet_grid(mr_sig ~ .)
```


Cis vs trans

```{r}
table(p.adjust(res_cis_coloc$p, "fdr") < 0.05)
table(p.adjust(res_trans$p, "fdr") < 0.05)
inner_join(res_cis_coloc, res_trans, by=c("id.exposure", "id.outcome")) %>%
{table(.$pval.x < 0.05, .$pval.y < 0.05)}

summary(lm(I(pval.x < 0.05) ~ I(pval.y < 0.05), data=inner_join(res_cis_coloc, res_trans, by=c("id.exposure", "id.outcome"))))
```

Are cis and trans MR results occurring more than expected by chance? For cis and trans MRs across protein pairs identify how many have p-values < alpha, then scramble cis and trans protein pairs 10000 times and determine if the observed number is larger than the distribution of null counts.

```{r}
nperm <- 10000
perm_res <- lapply(c(0.05, 0.01, 0.001, 0.05/nrow(temp)), function(thresh){
    temp <- inner_join(res_cis, res_trans, by=c("id.exposure", "id.outcome"))
    ncis <- sum(temp$pval.x < thresh)
    ntrans <- sum(temp$pval.y < thresh, na.rm=T)
    perms <- sapply(1:nperm, function(x) sum(sample(1:nrow(temp), ncis, replace=FALSE) %in% sample(1:nrow(temp), ntrans, replace=FALSE))) %>% sort()
    real <- inner_join(res_cis, res_trans, by=c("id.exposure", "id.outcome")) %>%
    {sum(.$pval.x < thresh & .$pval.y < thresh, na.rm=T)}
    tibble(thresh=thresh, nsig=real, maxperm=max(perms), medperm=median(perms), p_perm=sum(real < perms)/nperm)
}) %>% bind_rows()
perm_res
```

Cis and trans MR results co-occur more than expected by chance.

Do effect sizes correspond?

```{r}
inner_join(res_cis, res_trans, by=c("id.exposure", "id.outcome")) %>%
    filter(pval.x < 0.05) %>%
    ggplot(., aes(x=b.x, y=b.y)) +
    geom_point() +
    geom_smooth(method="lm") + 
    labs(x="cis-MR estimate", y="trans-MR estimate")
```

No concordance between cis and trans - just like in GoDMC! Restrict to cis coloc


```{r}
inner_join(res_cis_coloc, res_trans, by=c("id.exposure", "id.outcome")) %>%
    filter(pval.x < 0.05) %>%
    ggplot(., aes(x=b.x, y=b.y)) +
    geom_point() +
    geom_smooth(method="lm") + 
    labs(x="cis-MR estimate", y="trans-MR estimate")
```

```{r}
inner_join(res_cis_coloc, res_trans, by=c("id.exposure", "id.outcome")) %>%
    filter(pval.x < 0.05) %>%
    {summary(lm(b.y ~ b.x, .))}
```

```{r}
inner_join(res_cis, res_trans, by=c("id.exposure", "id.outcome")) %>%
    {summary(lm(b.y ~ b.x, .))}
```

```{r}
inner_join(res_cis, res_trans, by=c("id.exposure", "id.outcome")) %>%
    filter(pval.x < 0.05) %>%
    {summary(lm(b.y ~ b.x, .))}
```


## Forward vs reverse MR

Do forward and reverse MR results agree?

```{r}
bind_rows(res_cis %>% mutate(what="cis"), res_trans %>% mutate(what="trans")) %>%
    filter(!is.na(pval)) %>%
    inner_join(., prs_pairs, by=c("id.exposure"="prot", "id.outcome"="opengwasid")) %>%
    mutate(mr_sig = p.adjust(pval.x, "fdr") < 0.05) %>%
    ggplot(., aes(beta, b)) +
        geom_hline(yintercept=0, linetype="dotted") +
        geom_vline(xintercept=0, linetype="dotted") +
        geom_point(aes(colour=mr_sig)) +
        geom_smooth(method="lm") +
        labs(x="PRS disease -> protein", y="MR protein -> disease") +
        facet_grid(mr_sig ~ what)
```

Use steiger filtering. Note only trans instruments had any evidence of being reverse causal.

```{r}
table(dat$rsq.exposure > dat$rsq.outcome, dat$cistrans.exposure)
table(dat$steiger_pval < 0.05, dat$cistrans.exposure)
```

All reverse instruments are trans!

```{r}
bind_rows(
    res_cis %>% mutate(what="cis"), 
    res_cis_coloc %>% mutate(what="cis+coloc"), 
    res_trans %>% mutate(what="trans"), 
    res_trans_sf %>% mutate(what="trans sf"),
    res_trans_sf_relaxed %>% mutate(what="trans sf strict")) %>%
    filter(!is.na(pval)) %>%
    inner_join(., prs_pairs, by=c("id.exposure"="prot", "id.outcome"="opengwasid")) %>%
    mutate(mr_sig = p.adjust(pval.x, "fdr") < 0.05) %>%
    ggplot(., aes(beta, b)) +
        geom_hline(yintercept=0, linetype="dotted") +
        geom_vline(xintercept=0, linetype="dotted") +
        geom_point(aes(colour=mr_sig)) +
        geom_smooth(method="lm") +
        labs(x="PRS disease -> protein", y="MR protein -> disease") +
        facet_grid(mr_sig ~ what)
```


## PRS discovers forward MR?

```{r}
cis_prs <- subset(res_cis, paste(id.exposure, id.outcome) %in% paste(prs_pairs$prot, prs_pairs$opengwasid))
cis_prs$fdr <- p.adjust(cis_prs$pval, "fdr")
a <- subset(cis_prs, fdr < 0.05) %>% mutate(code = paste(id.exposure, id.outcome))
res_cis$fdr <- p.adjust(res_cis$pval, "fdr")
b <- subset(res_cis, fdr < 0.05) %>% mutate(code = paste(id.exposure, id.outcome))
sum(a$code %in% b$code)

tribble(~what, ~count,
    "Total tested - forward MR", nrow(res_cis),
    "Total discovered - forward MR", sum(res_cis$fdr < 0.05),
    "Total tested - reverse MR filter", nrow(cis_prs),
    "Total discovered - reverse MR filter", sum(cis_prs$fdr < 0.05),
    "Discovery overlap", sum(a$code %in% b$code)
)
```

Cis MR is substantially better powered than PRS for identifying forward MR assocs. But `r ((sum(cis_prs$fdr < 0.05) - sum(a$code %in% b$code)) / sum(res_cis$fdr < 0.05)) %>% {round(.*100, 1)}`% extra forward MR assocs discovered that wouldn't have been found by cis MR.

## Bidirectional effects?

Using cis for forward and any with !steiger for reverse

```{r}
rev_ids <- paste(dat$id.exposure[!dat$steiger_dir], dat$id.outcome[!dat$steiger_dir])
cis_mr_ids <- paste(res_cis_coloc$id.exposure, res_cis_coloc$id.outcome)
table(rev_ids %in% cis_mr_ids)
```

```{r}
bid <- subset(res_cis_coloc, cis_mr_ids %in% rev_ids)
bid
```


```{r}
rev_ids_prs <- paste(prs_pairs$prot, prs_pairs$opengwasid)
cis_mr_ids <- paste(res_cis_coloc$id.exposure, res_cis_coloc$id.outcome)
table(rev_ids_prs %in% cis_mr_ids)
```

```{r}
bidprs <- subset(res_cis_coloc, cis_mr_ids %in% rev_ids_prs)
bidprs
```


Some of these PRS results could just be driven by strong cis signals e.g. PCSK9.

```{r}
bidprs$ids <- paste(bidprs$id.exposure, bidprs$id.outcome)
bid_trans <- subset(res_trans, paste(id.exposure, id.outcome) %in% bidprs$ids) %>% mutate(fdr = p.adjust(pval, "fdr"))
bid_trans %>% subset(fdr < 0.05)
```

**Are these plausible bi-directional associations?**

## Perform instrument heterogeneity analysis

```{r}
het2 <- dat %>%
    subset(., paste(id.exposure, id.outcome) %in% paste(ressig$id.exposure, ressig$id.outcome)) %>%
    mr_heterogeneity(method_list="mr_ivw")
het2
```

For multi-instrument exposures do the MR using other methods, then test if the different methods are heterogeneous

```{r}
res2 <- dat %>%
    subset(., paste(id.exposure, id.outcome) %in% paste(ressig$id.exposure, ressig$id.outcome)) %>%
    mr()
fn <- function(b, se)
{
    rma.uni(b, sei=se, method="FE")$QEp
}
het_method <- res2 %>% filter(nsnp >= 3) %>% group_by(id.exposure, id.outcome) %>%
    do({
        x <- .
        tibble(method_Q_pval=fn(x$b, x$se))
    })
```

Plot comparison across different methods

```{r}
p <- res %>% 
    filter(nsnp > 2) %>% 
    filter(p.adjust(pval, "fdr") < 0.05) %>%
    mutate(code = paste(id.exposure, id.outcome)) %>% {.$code}
res2 %>% filter(paste(id.exposure, id.outcome) %in% p) %>%
ggplot(., aes(x=method, y=b)) +
geom_point(aes(colour=method)) +
geom_errorbar(aes(ymin=b-se*1.96, ymax=b+se*1.96, colour=method), width=0) +
facet_wrap(~ id.exposure+id.outcome, scale="free_y") 
```

For multi-instrument exposures do leave-one-out analysis to see if any are driven by a single SNP

```{r}
mr_leaveoneout2 <- function (dat, parameters = default_parameters(), method = mr_ivw) {
    if (!"samplesize.outcome" %in% names(dat)) {
        dat$samplesize.outcome <- NA
    }
    stopifnot("outcome" %in% names(dat))
    stopifnot("exposure" %in% names(dat))
    stopifnot("beta.exposure" %in% names(dat))
    stopifnot("beta.outcome" %in% names(dat))
    stopifnot("se.exposure" %in% names(dat))
    stopifnot("se.outcome" %in% names(dat))
    res <- plyr::ddply(dat, c("id.exposure", "id.outcome"), function(X) {
        x <- subset(X, mr_keep)
        nsnp <- nrow(x)
        if (nsnp == 0) {
            x <- X[1, ]
            d <- data.frame(SNP = "All", b = NA, se = NA, p = NA, 
                samplesize = NA, outcome = x$outcome[1], exposure = x$exposure[1])
            return(d)
        }
        if (nsnp > 2) {
            l <- lapply(1:nsnp, function(i) {
                with(x, method(beta.exposure[-i], beta.outcome[-i], 
                  se.exposure[-i], se.outcome[-i], parameters))
            })
            l[[nsnp + 1]] <- with(x, method(beta.exposure, beta.outcome, 
                se.exposure, se.outcome, parameters))
            d <- data.frame(SNP = c(as.character(x$SNP), "All"), 
                b = sapply(l, function(y) y$b), se = sapply(l, 
                  function(y) y$se), p = sapply(l, function(y) y$pval), 
                samplesize = x$samplesize.outcome[1])
            d$outcome <- x$outcome[1]
            d$exposure <- x$exposure[1]
        } else if(nsnp == 2) {
            l <- lapply(1:nsnp, function(i) {
                with(x, mr_wald_ratio(beta.exposure[-i], beta.outcome[-i], 
                  se.exposure[-i], se.outcome[-i], parameters))
            })
            l[[nsnp + 1]] <- with(x, method(beta.exposure, beta.outcome, 
                se.exposure, se.outcome, parameters))
            d <- data.frame(SNP = c(as.character(x$SNP), "All"), 
                b = sapply(l, function(y) y$b), se = sapply(l, 
                  function(y) y$se), p = sapply(l, function(y) y$pval), 
                samplesize = x$samplesize.outcome[1])
            d$outcome <- x$outcome[1]
            d$exposure <- x$exposure[1]

        }
        else {
            a <- with(x, method(beta.exposure, beta.outcome, 
                se.exposure, se.outcome, parameters))
            d <- data.frame(SNP = "All", b = a$b, se = a$se, 
                p = a$pval, samplesize = x$samplesize.outcome[1])
            d$outcome <- x$outcome[1]
            d$exposure <- x$exposure[1]
        }
        return(d)
    })
    res <- subset(res, select = c(exposure, outcome, id.exposure, 
        id.outcome, samplesize, SNP, b, se, p))
    return(res)
}
loo <- dat %>%
    subset(., paste(id.exposure, id.outcome) %in% paste(ressig$id.exposure, ressig$id.outcome)) %>%
    mr_leaveoneout2(.)
loo_passing <- loo %>% group_by(id.exposure, id.outcome) %>%
    filter(SNP != "All") %>%
    summarise(maxp=max(p, na.rm=T)) %>%
    filter(!is.na(maxp)) %>% 
    mutate(fdr=p.adjust(maxp, "fdr"))
```

Arrange all results

```{r}
allres <- left_join(
    ressig %>% select(id.exposure, id.outcome, nsnp, b, se, pval), 
    loo_passing %>% select(id.exposure, id.outcome, loo_maxp=maxp, loo_fdr=fdr)) %>% 
left_join(., het2 %>% select(id.exposure, id.outcome, inst_Q_pval=Q_pval)) %>%
left_join(., het_method) %>%
arrange(loo_maxp,  desc(inst_Q_pval)) %>% filter(!duplicated(paste(id.exposure, id.outcome)))
# subset(ressig, nsnp %in% 2:3) %>% 
#     select(id.exposure, id.outcome, nsnp, b, se, pval) %>% 
#     inner_join(., het2 %>% filter(method=="Inverse variance weighted") %>% select(id.exposure, id.outcome, inst_Q_pval=Q_pval)) %>% 
#     left_join(., loo_passing %>% select(id.exposure, id.outcome, loo_maxp=maxp)) %>%
#     arrange(loo_maxp, desc(inst_Q_pval))
# subset(ressig, nsnp==1) %>% arrange(pval)
traits <- readxl::read_xlsx(here("data", "Supplementary_Table5.xlsx")) %>% select(id.outcome=opengwasid, outcome=code)
m <- match(allres$id.outcome, traits$id.outcome)
all(traits$id.outcome[m] == allres$id.outcome)
allres$id.outcome <- traits$outcome[m]
allres
write.csv(allres, "forward_mr.csv")
```


## Are trans instruments more pleiotropic?

```{r}
dat$fdr.outcome <- p.adjust(dat$pval.outcome, "fdr")
table(dat$fdr < 0.05)

dat %>%
    group_by(SNP, cistrans.exposure) %>%
    filter(!duplicated(id.outcome)) %>%
    summarise(n=sum(fdr.outcome < 0.05)) %>%
    group_by(cistrans.exposure) %>%
    summarise(m=mean(n), sd=sd(n), med=median(n), min=min(n), max=max(n))
```

