---
title: Reverse MR analysis of PRS results
author: Gibran Hemani
execute:
  cache: true
---

```{r}
library(dplyr)
library(here)
library(TwoSampleMR)
library(tidyr)
library(ggplot2)

load(here("data", "all.rdata"))
prs_pairs <- readRDS(here("data", "prs_pairs.rds"))
dat <- readRDS(here("data", "reverse_mr_dat.rds"))
idinfo <- ieugwasr::gwasinfo(unique(dat$id.exposure))
keep_ids <- unique(dat$id.exposure)
keep_ids <- keep_ids[!keep_ids %in% c("ieu-a-299", "ieu-a-300", "ieu-a-301", "ieu-a-302", "ieu-b-38", "ieu-b-39")]
table(dat$id.exposure %in% keep_ids)
dat <- subset(dat, id.exposure %in% keep_ids)
dat$exposure[dat$id.exposure == "ieu-b-18"]
dat$exposure[dat$id.exposure == "ieu-a-805"]
subset(dat, is.na(exposure))$id.exposure %>% unique

# dat$id.exposure[dat$id.exposure == "ebi-a-GCST003374"] <- "ieu-a-1102"
# dat$id.exposure[dat$id.exposure == "ebi-a-GCST005047"] <- "ieu-a-26"
dat$id.exposure[dat$id.exposure == "ieu-a-1024"] <- "ieu-b-18"
dat$id.exposure[dat$id.exposure == "ieu-a-1187"] <- "ieu-a-805"
dat$exposure[dat$id.exposure == "ieu-a-805"] <- "MDD"
dat$exposure[dat$id.exposure == "ieu-b-18"] <- "Multiple sclerosis"
# MS might be on the wrong effect allele
dat$beta.exposure[dat$id.exposure == "ieu-b-18"] <- dat$beta.exposure[dat$id.exposure == "ieu-b-18"] * -1

prs_pairs$opengwasid[prs_pairs$opengwasid == "ieu-a-1187"] <- "ieu-a-805"
prs_pairs$opengwasid[prs_pairs$opengwasid == "ieu-a-1024"] <- "ieu-b-18"
table(unique(dat$id.exposure) %in% prs_pairs$opengwasid)
table(is.na(dat$exposure))
```

Perform MR

```{r}
rres <- suppressMessages(mr(dat, method=c("mr_ivw", "mr_wald_ratio")))
rres$fdr <- p.adjust(rres$pval, "fdr")
saveRDS(rres, file=here("data", "reverse_mr.rds"))
table(rres$fdr < 0.05)
```

Compare against PRS

```{r}
temp <- inner_join(prs_pairs, rres, by=c("prot"="outcome", "opengwasid"="id.exposure"))
dim(temp)
ggplot(temp, aes(x=Estimate, y=b)) +
geom_point() +
geom_smooth(method="lm") +
labs(x="PRS estimate", y="MR estimate") + facet_wrap(~ Disease_Name, scale="free")
```

Good agreement between PRS and IVW results

PRS analysis significance threshold = 2.3e-6

Heterogeneity

```{r}
het <- mr_heterogeneity(dat, method=c("mr_ivw"))
het$fdr <- p.adjust(het$Q_pval, "fdr")
table(het$fdr < 0.05)
```

Some SNPs have no heterogeneity

```{r}
nohet <- subset(het, fdr > 0.05)
head(nohet)
```

Select SNPs that have no heterogeneity and a significant IVW assoc

```{r}
resnohet <- subset(rres, (paste(exposure, outcome) %in% paste(nohet$exposure, nohet$outcome) | nsnp == 1) & fdr < 0.05)
table(resnohet$exposure) %>% as.data.frame
dim(resnohet)
```

These assocs seem to be most consistent with being reverse causal

```{r}
resnohet <- left_join(resnohet, nohet %>% select(id.exposure, id.outcome, Q_pval), by=c("id.exposure", "id.outcome"))
head(resnohet)
dim(resnohet)
write.csv(resnohet, file=here("results", "resnohet.csv"))
```

```{r}
resnohetall <- subset(rres, (paste(exposure, outcome) %in% paste(nohet$exposure, nohet$outcome) | nsnp == 1))
table(resnohetall$exposure) %>% as.data.frame
dim(resnohetall)
write.csv(resnohetall, file=here("results", "resnohetall.csv"))
```

Analyse the ones with large heterogeneity

1. Find outliers
2. Re-estimate with outliers removed


```{r}
est_qj <- function(d) {
    if(nrow(d) < 2) return(NULL)
    r <- suppressMessages(mr(d, method="mr_ivw") %>% mutate(what="All"))
    bhat <- r$b
    ss <- suppressMessages(mr_singlesnp(d)) %>% filter(!grepl("All", SNP)) %>%
    mutate(w = 1/(se^2), qj = w * (bhat - b)^2, qj_pval = pchisq(qj, df=1, low=F), qj_fdr = p.adjust(qj_pval, "fdr"))
    
    # remove MHC and outliers
    d[! (d$chr.outcome == 6 & d$pos.outcome > 27000000 & d$pos.outcome < 35000000), ]
    d_noout <- d %>% filter(!SNP %in% subset(ss, p.adjust(qj_pval, "fdr") < 0.05)$SNP)
    
    r2 <- suppressMessages(mr(d_noout, method="mr_ivw")) %>% mutate(what="Outliers removed")
    return(list(qj=ss, r=bind_rows(r, r2)))
}
qj <- group_by(dat, paste(id.exposure, id.outcome)) %>% 
    group_map(~ est_qj(.))

qjs <- lapply(qj, \(x) x$qj) %>% bind_rows()
res_out <- lapply(qj, \(x) x$r) %>% bind_rows() %>% select(names(.)[1:10])
head(res_out)
dim(res_out)
saveRDS(qjs, file=here("data", "reverse_mr_qj.rds"))
saveRDS(res_out, file=here("data", "reverse_mr_outliers.rds"))
```

Results table

```{r}
res_out %>% select(exposure, id=id.exposure, outcome, nsnp, b, se, pval, what) %>% write.csv(., file=here("results", "reverse_mr_results.csv"))
```


Example where there is no heterogeneity

```{r}
dr <- subset(dat, id.exposure==resnohet$id.exposure[1] & id.outcome==resnohet$id.outcome[1])
drr <- suppressMessages(mr(dr))
mr_scatter_plot(drr, dr)
```

Example where PRS result is driven by heterogeneity

```{r}
temp <- arrange(het, fdr)
dr2 <- subset(dat, id.exposure==temp$id.exposure[1] & id.outcome==temp$id.outcome[1])
drr2 <- suppressMessages(mr(dr2))
mr_scatter_plot(drr2, dr2)
```

Outlier analysis. How many have outliers but after removing them the original assoc still remains

```{r}
thresh <- subset(rres, fdr < 0.05) %>% arrange(desc(pval)) %>% {max(.$pval)}
res_out_sum <- res_out %>% group_by(id.exposure, id.outcome) %>%
    summarise(
        nsnp_orig = nsnp[1],
        noutliers = nsnp[1] - nsnp[2],
        fdr_orig = pval[1] < thresh,
        fdr_out = pval[2] < thresh,
        sign_orig = sign(b[1]),
        sign_out = sign(b[2]),
    ) %>% filter(fdr_orig)
head(res_out_sum)
sum(res_out_sum$fdr_orig & res_out_sum$fdr_out, na.rm=T)
```

Reverse MR summary

```{r}
tribble(
    ~what, ~n,
    "Number of PRS results", nrow(prs_pairs),
    "Number of reverse MR assocs", nrow(rres),
    "Reverse MR FDR < 0.05", sum(rres$fdr < 0.05, na.rm=T),
    "Reverse MR FDR < 0.05 and no heterogeneity", nrow(resnohet),
    "Reverse MR FDR < 0.05 after outlier removal", sum(res_out_sum$fdr_orig & res_out_sum$fdr_out, na.rm=T),
    "Mean number of instruments", mean(res_out_sum$nsnp_orig),
    "Mean number of outliers", mean(res_out_sum$noutliers, na.rm=T)
)

```

Compare PRS and reverse MR

```{r}
revmrpairs <- subset(res_out_sum, fdr_orig & fdr_out)
dim(revmrpairs)
rmp <- subset(res_out, what=="Outliers removed" & paste(id.exposure, id.outcome) %in% paste(revmrpairs$id.exposure, revmrpairs$id.outcome))
head(rmp)

temp <- left_join(prs_pairs, rres, by=c("prot"="outcome", "opengwasid"="id.exposure"))
temp$repl <- paste(temp$prot, temp$opengwasid) %in% paste(rmp$id.outcome, rmp$id.exposure)

head(temp)
table(temp$repl)
temp %>% filter(b < 1 & P_value < 0.05/nrow(temp)) %>%
    arrange(repl) %>%
ggplot(., aes(x=Estimate, y=b)) +
geom_point() +
geom_smooth(method="lm") +
geom_abline(slope=1, intercept=0, linetype="dotted") +
labs(x="PRS estimate", y="MR estimate", colour="MR FDR < 0.05")
ggsave(here("images", "mr_prs_comparison.pdf"))
```


1. sig in PRS, sig in revMR
    - Sensitive to outlier removal?
        - YES: remove from PRS
        - NO: keep in PRS
2. sig in PRS, not sig in revMR
    - Power is lower - mostly this
    - Pleiotropy is different
3. data unavailable - occasions possible

Overall

1. reverse MR vs PRS heterogeneity - Are they significantly different estimates
2. reverse MR outliers vs PRS heterogeneity - Are they significantly different estimates
3. drop =
    - revMR out vs PRS heterogeneity sig - fdr 
    - revMR out estimate overlaps the null



```{r}
qstat <- function(beta, se) {
    w <- 1/se^2
    bhat <- sum(beta * w) / sum(w)
    qj <- w * (beta - bhat)^2
    Q <- sum(qj)
    Qdf <- length(beta) - 1
    Qpval <- pchisq(Q, df=Qdf, lower.tail=FALSE)
    return(Qpval)
}

qstat(c(0.1, 0.1), c(0.1, 0.1))
qstat(c(0.1, -0.1), c(0.01, 0.01))

temp <- group_by(temp, prot, opengwasid) %>% mutate(Qpval = qstat(c(Estimate, b), c(Std_Error, se)))
temp

ggplot(temp %>% filter(P_value < 2.3e-6), aes(x=Estimate, y=b)) +
geom_point(aes(colour=Disease_Name)) +
geom_smooth(method="lm") +
facet_grid(. ~ I(Qpval < 0.05 & pval > 0.05))

```


```{r}
temp2 <- left_join(prs_pairs, res_out %>% filter(what == "Outliers removed"), by=c("prot"="outcome", "opengwasid"="id.exposure"))
temp2 <- group_by(temp2, prot, opengwasid) %>% mutate(Qpval = qstat(c(Estimate, b), c(Std_Error, se)), Qfdr = p.adjust(Qpval, "fdr"))

ggplot(temp2 %>% filter(P_value < 2.3e-6), aes(x=Estimate, y=b)) +
geom_point(aes(colour=Disease_Name)) +
geom_smooth(method="lm") +
facet_grid(. ~ I(Qpval < 0.05 & pval > 0.05))

ind <- temp2$P_value < 2.3e-6
table(temp$Qpval[ind] < 0.05, temp$pval[ind] > 0.05)
table(temp2$Qpval[ind] < 0.05, temp2$pval[ind] > 0.05)

```


```{r}
temp2 %>% filter(P_value < 2.3e-6) %>%
group_by(Disease_Name) %>% summarise(n(), Qpval = sum(Qfdr < 0.05), revmr_notsig = sum(pval > 0.05), remove = sum(Qfdr < 0.05 & pval > 0.05))
```

Remove due to PRS vs RevMRout heterogeneity

```{r}
temp2 <- temp2 %>% filter(P_value < 2.3e-6) %>% mutate(remove = Qpval < 0.05 & pval > 0.05, code=paste(opengwasid, prot))
table(is.na(temp2$remove))
prs_remove1 <- temp2$code[temp2$remove]
```

need to remove pairs where outliers change significance

```{r}
prs_sig <- temp2 %>% filter(P_value < 2.3e-6) %>% mutate(code = paste(opengwasid, prot)) %>% {.$code}
res_out_summary <- res_out %>% 
    mutate(code = paste(id.exposure, id.outcome)) %>%
    filter(code %in% prs_sig) %>%
    group_by(id.exposure, id.outcome) %>% 
    summarise(
        Qpval = qstat(b, se), 
        Qfdr = p.adjust(Qpval, "fdr"), 
        check = what[1] == "All",
        sig = pval[1] < 0.05,
        lostsig = pval[1] < 0.05 & pval[2] > 0.05, 
        gainedsig = pval[1] > 0.05 & pval[2] < 0.05, 
        attenuated = abs(b[1]) > abs(b[2])
    ) %>% 
    ungroup()

res_out_summary %>%
    group_by(id.exposure) %>%
    summarise(
        n=n(), 
        check = all(check),
        nsig = sum(sig, na.rm=T),
        nQsig = sum(Qfdr < 0.05, na.rm=T), 
        Qpvalsig = sum(Qpval < 0.05, na.rm=T), 
        remove = sum(Qpval < 0.05 & attenuated, na.rm=T),
        lostsig = sum(lostsig, na.rm=T)
    )
```

```{r}
prs_temp <- prs_pairs %>% select(id.exposure = opengwasid, id.outcome = prot, b = Estimate, se= Std_Error, pval = P_value) %>% mutate(what = "PRS")
```


```{r}
res_out %>% bind_rows(., prs_temp) %>% filter(id.exposure == "ieu-b-18", paste(id.exposure, id.outcome) %in% prs_sig) %>%
    left_join(., res_out_summary) %>%
    ggplot(., aes(x=b, y=id.outcome)) +
    geom_point(aes(colour=what), position=ggstance::position_dodgev(height=0.4)) +
    geom_errorbarh(aes(colour=what, xmin=b-1.96*se, xmax=b+1.96*se), position=ggstance::position_dodgev(height=0.4), height=0) +
    scale_colour_brewer(type="qual") +
    facet_wrap(~ I(Qpval < 0.05), scale="free_y")
```

```{r}
res_out %>% bind_rows(., prs_temp) %>% filter(id.exposure == "ieu-b-18", paste(id.exposure, id.outcome) %in% prs_sig) %>%
    left_join(., res_out_summary) %>%
    ggplot(., aes(x=b, y=id.outcome)) +
    geom_point(aes(colour=what), position=ggstance::position_dodgev(height=0.4)) +
    geom_errorbarh(aes(colour=what, xmin=b-1.96*se, xmax=b+1.96*se), position=ggstance::position_dodgev(height=0.4), height=0) +
    scale_colour_brewer(type="qual") +
    facet_wrap(~ lostsig)
```

```{r}
res_out %>% bind_rows(., prs_temp) %>% filter(id.exposure == "ieu-b-18", paste(id.exposure, id.outcome) %in% prs_sig) %>%
    left_join(., res_out_summary) %>%
    ggplot(., aes(x=b, y=id.outcome)) +
    geom_point(aes(colour=what), position=ggstance::position_dodgev(height=0.4)) +
    geom_errorbarh(aes(colour=what, xmin=b-1.96*se, xmax=b+1.96*se), position=ggstance::position_dodgev(height=0.4), height=0) +
    scale_colour_brewer(type="qual") +
    facet_wrap(~ I(Qpval < 0.05 & attenuated))
```


- If it was PRS sig but not revMR sig, probably not an outlier driving it
- If it was PRS sig and revMR



```{r}
prs_remove2 <- res_out_summary %>% mutate(code=paste(id.exposure, id.outcome)) %>% filter(lostsig) %>% {.$code}
prs_remove <- c(prs_remove1, prs_remove2) %>% unique
```


```{r}
prs_pairs_sig <- subset(prs_pairs, P_value < 2.3e-6)
prs_pairs_sig$Disease_Name %>% unique

prs_pairs$Disease_Name %>% unique
nrow(prs_pairs_sig)
rres_prs_sig <- subset(rres, paste(id.exposure, id.outcome) %in% paste(prs_pairs_sig$opengwasid, prs_pairs_sig$prot))
dim(rres_prs_sig)
```


```{r}
rres_prs_sig$fdr <- p.adjust(rres_prs_sig$pval, "fdr")
table(rres_prs_sig$fdr < 0.05)
table(rres_prs_sig$pval < 0.05)
```


```{r}
prs_pairs_sig$code <- paste(prs_pairs_sig$opengwasid, prs_pairs_sig$prot)
rres_prs_sig$code <- paste(rres_prs_sig$id.exposure, rres_prs_sig$id.outcome)

tem <- inner_join(prs_pairs_sig, rres_prs_sig, by="code") %>% 
    mutate(prs_sign = sign(Estimate))
tem$Estimate <- abs(tem$Estimate)
tem$b[tem$prs_sign == -1] <- tem$b[tem$prs_sign == -1] * -1


ggplot(tem, aes(x=Estimate, y=b)) + geom_point(aes(colour=exposure)) +
geom_smooth(method="lm", aes(colour=fdr.y < 0.05)) +
geom_abline() +
facet_grid(. ~ I(fdr.y < 0.05)) + labs(x="PRS", y="RevMR")
```


```{r}
ggplot(tem, aes(x=Estimate, y=b)) + geom_point(aes(colour=exposure)) +
geom_smooth(method="lm", aes(colour=fdr.y < 0.05)) +
geom_abline() +
facet_grid(. ~ I(code %in% prs_remove)) + labs(x="PRS", y="RevMR")
```

Organise results

- PRS results
    - id.exposure
    - exposure
    - outcome
    - beta
    - se
    - pval
    - Qpval
    - remove

- RevMR results
    - id.exposure
    - exposure
    - outcome
    - beta
    - se
    - pval
    - nsnp
    - noutliers
    - Qpval
    - beta_out
    - se_out
    - pval_out


```{r}
prs_organised <- temp2 %>% select(exposure = Disease_Name, id.exposure = opengwasid, outcome = prot, prs_beta = Estimate, prs_se = Std_Error, prs_pval = P_value, prs_Qpval_revmrout = Qpval) %>% mutate(remove = paste(id.exposure, outcome) %in% prs_remove)
```

```{r}
revmr_organised <- res_out %>% 
    filter(what == "All", paste(id.exposure, id.outcome) %in% prs_sig) %>% 
    select(id.exposure = id.exposure, outcome = outcome, revmr_nsnp = nsnp, revmr_beta = b, revmr_se = se, revmr_pval = pval)
revmr_organised <- left_join(revmr_organised, res_out %>% 
    filter(what == "Outliers removed", paste(id.exposure, id.outcome) %in% prs_sig) %>% 
    select(id.exposure = id.exposure, outcome = outcome, revmr_nsnp_out = nsnp, revmr_beta_out = b, revmr_se_out = se, revmr_pval_out = pval), by=c("id.exposure", "outcome")) %>% filter(!duplicated(paste(exposure, outcome)))
revmr_organised <- left_join(revmr_organised, res_out_summary %>% select(id.exposure, outcome=id.outcome, revmr_Qpval_revmrout = Qpval))

head(revmr_organised)
```

```{r}
rev_organised <- left_join(prs_organised, revmr_organised, by=c("id.exposure", "outcome"))
rev_organised$remove <- rev_organised$remove | is.na(rev_organised$revmr_beta) | is.na(rev_organised$prs_Qpval_revmrout) | is.na(rev_organised$revmr_Qpval_revmrout)
rev_organised
table(is.na(rev_organised$revmr_beta))
table(rev_organised$remove)
write.csv(rev_organised, here("results", "revmr_organised.csv"))
```




Anything not available in the reverse MR, set PRS to 0? (n=9)


Table -
- PRS sig
- reverse MR
    - flag - outlier removal leads to heterogeneity with PRS
    - flag = PRS unreliable
- forward MR


ieu-a-805 = MDD
ieu-b-18 = MS - might have the wrong effect allele




Plot rev MR against rev MR outliers

```{r}

inner_join(res_out %>% filter(what == "All"), res_out %>% filter(what == "Outliers removed"), by=c("id.exposure", "id.outcome")) %>% ggplot(., aes(b.x, b.y)) + geom_point(aes(colour=exposure.x)) + geom_abline(a=0, b=1)


```



```{r}
subset(temp, Estimate < 0 & b > 0 & repl)
```


```{r}
rownames(rmp) <- NULL
write.csv(rmp, file=here("results", "resnohet.csv"))
```

Other Summary

```{r}
cis_prs <- subset(res_cis, paste(id.exposure, id.outcome) %in% paste(prs_pairs$prot, prs_pairs$opengwasid))
cis_prs$fdr <- p.adjust(cis_prs$pval, "fdr")
a <- subset(cis_prs, fdr < 0.05) %>% mutate(code = paste(id.exposure, id.outcome))
res_cis$fdr <- p.adjust(res_cis$pval, "fdr")
b <- subset(res_cis, fdr < 0.05) %>% mutate(code = paste(id.exposure, id.outcome))
sum(a$code %in% b$code)

colocres <- readRDS(here("data", "cis_coloc_res.rds"))
res_cis$fdr <- p.adjust(res_cis$pval, "fdr")
tribble(
    ~what, ~n,
    "Forward MR tests", nrow(res),
    "Forward MR tests using cis-only instruments", nrow(res_cis),
    "Forward MR tests using trans-only instruments", nrow(res_trans),
    "Cis MR with FDR < 0.05", sum(res_cis$fdr < 0.05),
    "Total tested - reverse MR filter", nrow(cis_prs),
    "Total discovered - reverse MR filter", sum(cis_prs$fdr < 0.05),
    "Discovery overlap", sum(a$code %in% b$code),
    "Cis Colocalisation with H4 > 0.8", sum(colocres$PP.H4.abf > 0.8),
)
```

How many prs_pairs per trait

```{r}
table(prs_pairs$opengwasid) %>% as.data.frame() %>% arrange(Freq) %>% left_join(., traits %>% select(code, opengwasid), by=c("Var1"="opengwasid"))
```

Examine CKD - it has a lot of assocs

```{r}
ckd <- subset(dat, exposure == "CKD")
ckd %>% group_by(id.outcome) %>%
    arrange(pval.outcome) %>%
    slice_head(n=1) %>%
    {table(.$SNP)}
```

Most of the time the best SNP is a single SNP. Heterogeneity?

```{r}
qjs %>% filter(exposure == "CKD") %>%
    group_by(SNP) %>%
    summarise(mean(qj))
```

Only 4 exposure SNPs.. 

```{r}
subset(ckd, outcome == outcome[1]) %>% select(SNP, pval.exposure)
```

Plot an example

```{r}
# note that prs_pairs uses different opengwasid
ckd_top <- subset(ckd, outcome %in% subset(prs_pairs, opengwasid == "ieu-a-1102")$prot[1:28])

ggplot(ckd_top, aes(x=beta.exposure, y=beta.outcome)) +
geom_point() +
facet_wrap(~id.outcome) +
geom_smooth(method="lm")
```

The pattern of association is identical across all these proteins. I don't know what to make of this. The 4 SNPs for CKD have a consistent effect on loads of blood protein levels?

Look at lung cancer too?

```{r}
sclc <- subset(dat, id.exposure == "ieu-a-989" & id.outcome %in% subset(res_out_sum, id.exposure=="ieu-a-989")$id.outcome)

sclc_top <- subset(dat, id.exposure == "ieu-a-989" & id.outcome %in% subset(prs_pairs, opengwasid=="ieu-a-989")$prot[1:28])

ggplot(sclc, aes(x=beta.exposure, y=beta.outcome)) +
geom_point() +
facet_wrap(~id.outcome) +
geom_smooth(method="lm")
```

Plot MR against PRS

```{r}
qjs %>% filter(id.exposure=="ieu-a-989") %>%
    group_by(SNP) %>%
    summarise(mean(qj))
```

```{r}
res_out_sum %>%
    group_by(id.exposure) %>%
    summarise(fdr_out=sum(fdr_out, na.rm=T), fdr_orig=sum(fdr_orig, na.rm=T)) %>%
    left_join(., traits %>% select(id.exposure=opengwasid, code)) %>% 
    left_join(., prs_pairs %>% group_by(id.exposure=opengwasid) %>% summarise(nprs=n())) %>%
    arrange(desc(nprs)) %>%
    select(code, id.exposure, n_prs=nprs, n_mr=fdr_orig, n_mr_out=fdr_out)
```

T2D

```{r}
# note that prs_pairs uses different opengwasid
t2d_top <- subset(dat, id.exposure == "ieu-a-26" & id.outcome %in% unique(subset(prs_pairs, opengwasid=="ieu-a-26"))$prot[1:56])

ggplot(t2d_top, aes(x=beta.exposure, y=beta.outcome)) +
geom_point() +
facet_wrap(~id.outcome) +
geom_smooth(method="lm")
```

MS

```{r}
ms_top <- subset(dat, id.exposure == "ieu-b-18" & id.outcome %in% unique(subset(prs_pairs, opengwasid=="ieu-b-18"))$prot[1:56])

ggplot(ms_top, aes(x=beta.exposure, y=beta.outcome)) +
geom_point() +
facet_wrap(~id.outcome, scale="free_y") +
geom_smooth(method="lm") +
theme(axis.text.x=element_blank(), axis.text.y=element_blank())
```


```{r}
plot_rev_mr <- function(id, nplot) {
    ms_top <- subset(dat, id.exposure %in% id & id.outcome %in% unique(subset(prs_pairs, opengwasid %in% id))$prot[1:nplot])
    ind <- sign(ms_top$beta.exposure) == -1
    ms_top$beta.exposure <- abs(ms_top$beta.exposure)
    ms_top$beta.outcome[ind] <- ms_top$beta.outcome[ind] * -1
    ms_top <- subset(ms_top, ! (chr.outcome == 6 & pos.outcome > 28000000 & pos.outcome < 33000000))

    ggplot(ms_top, aes(x=beta.exposure, y=beta.outcome)) +
    geom_point() +
    facet_wrap(~id.outcome, scale="free_y") +
    geom_smooth(method="lm") +
    theme(axis.text.x=element_blank(), axis.text.y=element_blank())
}

# plot_rev_mr("ebi-a-GCST003374", 56)
```

MS excluding chr 6

```{r}
plot_rev_mr(c("ieu-a-1024", "ieu-b-18"), 56)
```

```{r}
subset(dat, exposure == "MS") %>% filter(outcome == outcome[1])
```
